---
title: "Correlates of ATS"
author: "John Williams"
output: 
  Grmd::docx_document:
    fig_caption: TRUE
    force_captions: TRUE
    toc: TRUE
  pdf_document:
    toc: true
    latex_engine: xelatex
    keep_tex: true
---

```{r label=R-setup, echo=FALSE, include=FALSE, cache=FALSE}
# bibliography: /home/john/Dropbox/Writing/bib/all-refs.bib
require( ggplot2 )
require( scales )
require( Gmisc )
require( rms )
require( knitr )

# Evaluate the figure caption after the plot
knitr::opts_knit$set(eval.after='fig.cap')

# Use the table counter that the htmlTable() provides
options(table_counter = TRUE)

# Use the figCapNo() with roman letters
options(fig_caption_no_roman = TRUE)

#theme_set( theme_gray( base_size = 10 ))
#theme_update( legend.key.width = unit( 3,"line") )
options(width=120)
options("show.signif.stars"=F)

## set global chunk options
opts_chunk$set(echo=FALSE,
               cache=TRUE,
               dpi=96,
               fig.height=6,
               fig.width=8,
               dev="png",
               dev.args=list(type="cairo"),
               fig.align='center',
               fig.show='hold',
               par=TRUE,
               comment=NA,
               background="wheat",
               prompt=TRUE,
               warning=FALSE,
               message=FALSE)

info   <- sessionInfo()
r_ver  <- paste(info$R.version$major, info$R.version$minor, sep=".")
barcol <- "wheat"
source("/home/john/Dropbox/Writing/Software//R/logistic-regression-functions.R")
```

# Correlates of Active Transport to School

## Data Setup
```{r label="read-data", results='hide', echo=TRUE}
require(foreign)
dir  <- "/home/john/Dropbox/Research/Collaboration/BEATS/John/Data"
file <- "BEATS_StudentS_DATA-All_150424.sav"
dat <- read.spss( paste(dir, file, sep="/"), to.data.frame=TRUE)
```

```{r label=data-setup, results='hide', cache=TRUE}
require( car )
dat$ATS_n      <- recode( as.numeric(dat$ATvsMT), "1=0; 2=1") 
dat$whodecides <- recode(dat$tsdecision, "'My parents'='Other(s)'; 'School'='Other(s)'")
detach(package:car)
dat$ATS_f <- factor( dat$ATS_n, labels=c("Motorised", "Active") )
levels(dat$NZDepCat3) <- c("High", "Med", "Low")

# Transform rating scale variables to numeric
dat$tired         <- as.numeric(dat$WStired)
dat$confd         <- as.numeric(dat$WSconf)
dat$pshould       <- as.numeric(dat$WSpsh)
dat$onway         <- as.numeric(dat$WCSone)
dat$punsafe       <- as.numeric(dat$WSpunsafe)
dat$chat          <- as.numeric(dat$WSchat)
dat$time          <- as.numeric(dat$WStime)
dat$sched         <- as.numeric(dat$WSbsched)
dat$control       <- as.numeric(dat$WScontrol)
dat$pwalk         <- as.numeric(dat$WSpwalk)
dat$n_cars        <- as.numeric(dat$HMcars)
dat$regwalk       <- as.numeric(dat$TgWalk)

label(dat$tired) <- "Often too tired"

# create ordered factors and drop unused levels in the process
dat$school_decile   <- ordered(   dat$ScDecile)
dat$school_decile_n <- as.numeric(dat$school_decile)

# Make more print-friendly labels and levels
levels(dat$HMcars)[5] <- "Four+"
levels(dat$PAGuideQ)  <-c("No", "Yes")

dat.ats <- dat[complete.cases(dat$ATvsMT, dat$Dist2School, dat$Age_Cat, 
                          dat$aGender, dat$HMcars, dat$TgWalk,
                          dat$tired, dat$confd, dat$pshould, dat$punsafe, dat$chat, dat$time, dat$onway,
                          dat$TgWalk) & dat$boarder2 == "Not a boarder", ]

dat.full <- dat[complete.cases(dat$Dist2School, dat$ATS_f), c("Dist2School", "ATS_f") ]
cutoff <- 4800
dat.ats <- dat.ats[ dat.ats$Dist2School <= cutoff, ]

# Set up stuff for RMS functions
attach(dat.ats)
m1.ddist  <- datadist(ATS_n, Dist2School, regwalk, n_cars, tired, confd, pshould, onway, punsafe, chat, time, sched, control, pwalk, whodecides )
detach(dat.ats)
options(datadist="m1.ddist")
```

## Introduction

This document contains preliminary analyses of correlates of ATS (**A**ctive 
**T**ransport to **S**chool). We know from previous work (and common sense!) 
that the most influential correlate is distance from school. But once that is 
factored out, what else is influential?

```{r label=ATS-Dist, dependson="data-setup", fig.cap="Figure 1: Empirical probability of ATS by distance from school"}
emp <- NULL
with( dat.full,
      {
        for ( i in 1:length(Dist2School ) ) 
          emp[i] <- prop.table(table(ATS_f[ Dist2School < Dist2School[i]]))["Active"]
        emp_prob <- data.frame( dist=Dist2School, true=ATS_f, prob=emp )
        p <- ggplot( emp_prob, aes( dist, prob) ) 
        p <- p + labs(x="Distance from School (m)", y="Probability of ATS", 
                      title="Empirical probability of ATS")
        p <- p + geom_point(size=1) 
        p + geom_smooth(size=1)
        }
      )
```

The figure above shows the empirical probability of ATS, i.e. the proportion of 
respondents who use ATS at each level of observed distance to school. The observations 
plotted are those respondents who live less than 8km from school, however the analysis
below is restricted to students who live less than `r I(cutoff)`m from school.

## Descriptive statistics

Firstly, let's look at the distributions of the variables we'll be dealing with. 
The starting model will be the same model as previously used for the 
*Journal of Science and Medicine in Sport* manuscript, so we need to assess 
the distribution of the variables used in that model.

The analyses below are restricted to the students who live less than 
`r I(cutoff)`m from school, and who are not boarders. There are
`r I(dim(dat.ats)[1])` students who fit that criterion.

```{r label="descriptives", dependson="data-setup", fig.cap="Table 1"}
attach( dat.ats )
HTML <- TRUE
d_age <- getDescriptionStatsBy( Age_Cat,         ATS_f, html=HTML, show_all_values=TRUE, statistics=TRUE)
d_BMI <- getDescriptionStatsBy( BMI_2cat,        ATS_f, html=HTML, show_all_values=TRUE, statistics=TRUE)
d_bik <- getDescriptionStatsBy( HMbike,          ATS_f, html=HTML, show_all_values=TRUE, statistics=TRUE)
d_car <- getDescriptionStatsBy( HMcars,          ATS_f, html=HTML, show_all_values=TRUE, statistics=TRUE)
d_con <- getDescriptionStatsBy( confd,           ATS_f, html=HTML, show_all_values=TRUE, statistics=TRUE)
d_dec <- getDescriptionStatsBy( tsdecision,      ATS_f, html=HTML, show_all_values=TRUE, statistics=TRUE)
d_dep <- getDescriptionStatsBy( NZDepCat3,       ATS_f, html=HTML, show_all_values=TRUE, statistics=TRUE)
d_dis <- getDescriptionStatsBy( Dist2School,     ATS_f, html=HTML, show_all_values=TRUE, statistics=TRUE, header_count=TRUE )
d_onw <- getDescriptionStatsBy( onway,           ATS_f, html=HTML, show_all_values=TRUE, statistics=TRUE)
d_pag <- getDescriptionStatsBy( PAGuideQ,        ATS_f, html=HTML, show_all_values=TRUE, statistics=TRUE)
d_pun <- getDescriptionStatsBy( punsafe,         ATS_f, html=HTML, show_all_values=TRUE, statistics=TRUE)
d_psh <- getDescriptionStatsBy( pshould,         ATS_f, html=HTML, show_all_values=TRUE, statistics=TRUE)
d_reg <- getDescriptionStatsBy( regwalk,         ATS_f, html=HTML, show_all_values=TRUE, statistics=TRUE)
d_scd <- getDescriptionStatsBy( school_decile_n, ATS_f, html=HTML, show_all_values=TRUE, statistics=TRUE)
d_scr <- getDescriptionStatsBy( ScrGuide,        ATS_f, html=HTML, show_all_values=TRUE, statistics=TRUE)
d_sex <- getDescriptionStatsBy( aGender,         ATS_f, html=HTML, show_all_values=TRUE, statistics=TRUE)
d_sib <- getDescriptionStatsBy( hsiblings,       ATS_f, html=HTML, show_all_values=TRUE, statistics=TRUE)
d_tir <- getDescriptionStatsBy( tired,           ATS_f, html=HTML, show_all_values=TRUE, statistics=TRUE)
d_tim <- getDescriptionStatsBy( time,            ATS_f, html=HTML, show_all_values=TRUE, statistics=TRUE)

label(d_dis) <- "Distance to school (m)"
label(d_age) <- "Age"
label(d_sex) <- "Sex"
label(d_BMI) <- "BMI"
label(d_car) <- "Number of cars at home"
label(d_bik) <- "Number of bikes available for ATS at home"
label(d_sib) <- "Number of siblings"
label(d_dep) <- "NZ Deprivation Index"
label(d_tir) <- "Too tired to use ATS"
label(d_psh) <- "Parents say I should use ATS"
label(d_onw) <- "School is on the way to work"
label(d_scr) <- "Meets screen time guidelines"
label(d_pun) <- "My parents think its unsafe to use ATS"
label(d_reg) <- "Travel in general: Walking"
label(d_tim) <- "It takes too much time to walk to school"
label(d_dec) <- "Who makes the decision on travel mode?"
label(d_scd) <- "School decile (1--6=least deprived)"
label(d_pag) <- "Physical activity guidelines"

rgroup.2   <- c(label(d_dis), label(d_age), label(d_sex), label(d_BMI), 
                label(d_car), label(d_dep), label(d_tir), label(d_psh), 
                label(d_onw), label(d_scr), label(d_pun), label(d_reg), 
                label(d_tim), label(d_sib), label(d_bik), label(d_dec), 
                label(d_scd), label(d_pag))

n.rgroup.2 <- c( NROW(d_dis),  NROW(d_age),  NROW(d_sex),  NROW(d_BMI),  
                 NROW(d_car),  NROW(d_dep),  NROW(d_tir),  NROW(d_psh),  
                 NROW(d_onw),  NROW(d_scr),  NROW(d_pun),  NROW(d_reg),  
                 NROW(d_tim),  NROW(d_sib),  NROW(d_bik),  NROW(d_dec),
                 NROW(d_scd),  NROW(d_pag))

rgroup.1   <- c("Personal", "Household")
n.rgroup.1 <- c(4, 3)
  
my_rgroup   <- rbind(  rgroup.1,   rgroup.2)
my_n.rgroup <- rbind(n.rgroup.1, n.rgroup.2)
options(table_counter=TRUE)
options(table_counter_str="<strong>Table %s</strong>:")
htmlTable(x = rbind( d_dis, d_age, d_sex, d_BMI, 
                     d_car, d_dep, d_tir, d_psh, 
                     d_onw, d_scr, d_pun, d_reg,
                     d_tim, d_sib, d_bik, d_dec,
                     d_scd, d_pag),
  rgroup   = rgroup.2,
  n.rgroup = n.rgroup.2,
  caption  = paste("Individual and household potential correlates of ATS"),
  tfoot    = "<span style='font-size: 80%'>&dagger;All continuous variables are reported with mean and standard deviation: x&#772; (SD). Categorical variables are reported in counts and percentages: count (%).</br>&ddagger; The <i>p</i>-values are from Fisher tests for categorical variables and Wilcoxon tests for continuous variables.</span>",
  label    = "Table1",
  rowlabel = "Variable<sup>&dagger;</sup>",
  ctable   = TRUE)
detach( dat.ats )
```

### Plots

```{r label=plot-dist, fig.cap = "Figure 2: histogram of distance to school (bin width=100m)", dependson="data-setup"}
ggplot(dat.ats, aes(x=Dist2School)) + geom_histogram( binwidth=100, fill=barcol ) + labs(x=NULL, title="Distance to school (m)")
```

```{r label=plots-age, fig.cap = "Figure 3: bar plot of cars", dependson="data-setup"}
ggplot(dat.ats, aes(x=HMcars )) + geom_bar(fill=barcol) + labs( x=NULL, title="Home: n of cars, vans, trucks or utes")
```

```{r label=plots-tired, fig.cap="Figure 4: bar plot of tired", dependson="data-setup"}
ggplot(dat.ats, aes(x=WStired))+geom_bar(fill=barcol) + labs( x=NULL, title="I often feel too tired to walk to school") + theme(axis.text.x=element_text(angle=45, hjust=1) )
```

```{r label=plots-onway, fig.cap="Figure 5: bar plot of onway", dependson="data-setup"}
ggplot(dat.ats, aes(x=WCSone ))+geom_bar(fill=barcol) +labs(x=NULL, title=" It is easier for someone to drive me to school,\non the way to something else ")+ theme(axis.text.x=element_text(angle=45, hjust=1))
```

```{r label=plots-sched, fig.cap="Figure 6: bar plot of sched", dependson="data-setup"}
ggplot(dat.ats, aes(x=WSbsched  ))+geom_bar(fill=barcol) +labs(x=NULL, title="It is not convenient for me to walk to school\nbecause of my after-school schedule")+ theme(axis.text.x=element_text(angle=45, hjust=1))
```

```{r label=plots-contr, fig.cap="Figure 7: bar plot of contr", dependson="data-setup"}
ggplot(dat.ats, aes(x=WScontrol))+geom_bar(fill=barcol) +labs(x=NULL, title="How much personal control do you yourself\nhave over whether or not you walk to school?")+ theme(axis.text.x=element_text(angle=45, hjust=1))
```

```{r label=plots-confd, fig.cap="Figure 8: bar plot of confd", dependson="data-setup"}
ggplot(dat.ats, aes(x=WSconf  ))+geom_bar(fill=barcol) +labs(x=NULL, title="How confident are you that you could walk to school?")+ theme(axis.text.x=element_text(angle=45, hjust=1))
```

```{r label=plots-pshld, fig.cap="Figure 9: bar plot of pshld", dependson="data-setup"}
ggplot(dat.ats, aes(x=WSpsh  ))+geom_bar(fill=barcol) +labs(x=NULL, title="My parents or guardians think I should walk to school")+ theme(axis.text.x=element_text(angle=45, hjust=1))
```

```{r label=plots-regwalk , fig.cap="Figure 10: bar plot of regwalk ", dependson="data-setup"}
ggplot(dat.ats, aes(x=TgWalk))+geom_bar(fill=barcol) +labs(x=NULL, title="Travel in general: Walking")+ theme(axis.text.x=element_text(angle=45, hjust=1))
```


## Modeling

### Model 0
The most influential variable influencing active transport to school is distance 
to school. Hence, a logistic regression model with distance as the single 
explanantory variable will be the "null" or baseline model for model development
purposes. However, as we can see from Figure 1, the relationship between distance 
and probagility of ATS is clearly nonlinear. So a null model should also accomodate 
this fact. I chose to accomodate nonlinearity by use of a restricted cubic spline with 
three knots.


```{r label=ATS-M0, dependson="data-setup", results='asis', cache=TRUE} 
options(knots=3, poly.degree=2)
m0.dat <- dat.ats[complete.cases(dat.ats$ATS_f, dat.ats$Dist2School), 
                  c("ATS_n", "ATS_f", "Dist2School") ]
m0.ddist <- datadist(m0.dat)
options( datadist = "m0.ddist" )
m0     <- lrm( ATS_f ~ rcs(Dist2School,3), data=m0.dat, x=T, y=T ) 
m0.gof <- LR.summary( m0, html = TRUE )
```

We can see that this model is reasonably good, for example the area under the 
ROC curve is `r I(m0.gof$ROC)`%, and the Nagelkerke *R^2* is 
`r I(m0.gof$Nagelkerke)`%. Also there are only a few influential observations.



The summary table below shows the estimates and 95% confidence intervals for both
the logit parameters and ORs. These effects are computed over the IQR of continuous variables
as opposed to a 1-unit range, as is common with most statistical software.  

```{r ATS-MO-summary, dependson=c("data-setup", "ATS-M0"), results='asis'}
options(datadist="m1.ddist")
myHTML( summary( m0 ) )
```

For comparison, the OR for distance, computed over a 1-unit (i.e. 1m) range, is
`r I(round(exp(m0$coefficients)["Dist2School"], 4))`.  Because the units involved in 
subsequent analyses are quite different, the IQR method allows a type of 
standardisation, and hence is more intepretable. This parallels the use of betas and
standardised betas in OLS regression.

The ANOVA below shows likelihood ratio tests for all the parameter estimates. In
this case, the *p*-values are indentical with those in the table above, because the
single explanatory variable is continuous. However when an explanatory variable is
categorical, the logit model estimates a separate *p*-value for each dummy term,
while the ANOVA table inlcudes only a single *p*-value for all levels of the factor.

```{r ATS-MO-anova, dependson="ATS-M0", results='asis'}
myHTML(anova( m0 ))
```

When iteratively developing a model based in a single data set, it is well known
that there is a severe danger of "over-fitting", i.e. developing a model that
fits one data set well, but may not fit another random sample drawn from the 
sample population.  Fortunately, methods exist to estimate the degree of 
over-fitting, based on re-sampling (i.e. bootstrapping). This approach is known
as validating a model.
 
The validation results printed below show bias-corrected estimates of the 
discriminant and predictive accuracy indices. We can see that these results
indicate low optimism, i.e. the original indices are quite similar to the
bias-corrected estimates.


```{r ATS-MO-validate, dependson="ATS-M0",results='asis'}
samples <- 200
myHTML( validate( m0, B=samples ) )
```

The results above are based on `r I(samples)` bootstrap samples, as shown in the
rightmost column. We can see that in this case, the original and bias-corrected
fit indices are very similar, so we can be relatively confident that these results
are not specific to the particular sample they were calculated from, i.e. they 
are generalisable.

### Model 1
Now lets use the model from previously submitted manuscripts.

The New Zealand Deprivation Index categories have been collapsed into three, and
the baseline category is the lowest category (i.e highest deprivation).
```{r label=ATS-M1, dependson="data-setup"} 
attach( dat.ats )
dat.m1 <- dat.ats[complete.cases(ATS_n, Dist2School, n_cars, tired, confd, pshould, 
                             onway, punsafe, time, sched, control, pwalk, 
                             regwalk, whodecides), 
              c("ATS_f", "Dist2School", "n_cars", "tired", "confd", "pshould", 
                "onway", "punsafe", "time", "sched", "control", "pwalk", 
                "regwalk", "whodecides")]
detach( dat.ats )
attach( dat.m1 )
      m1.ddist  <- datadist(ATS_f, Dist2School, n_cars, tired, confd, pshould, 
                            onway, punsafe, time, sched, control, pwalk, 
                            regwalk, whodecides )
detach( dat.m1 )
options(datadist="m1.ddist")
m1 <- lrm( ATS_f ~ rcs(Dist2School, 3) + n_cars + tired + onway + confd + 
             pshould, x=T, y=T, data=dat.m1) 
```

The fuller model has a much higher area under the ROC curve.

```{r label=ATS-M1-print, dependson="ATS-M1",results='asis'} 
options(datadist='m1.ddist')
m1.gof  <- LR.summary( m1, html = TRUE) 
```

The effects of distance, more cars available to the household, being tired, 
and it being easier for someone to drive the student to school on the way 
to something else are negative, and the effects of being confident, having the 
encouragement of parents and parental encouragement are positive.

```{r label=ATS-M1-summary, dependson=c("data-setup", "ATS-M1"), results='asis'} 
options(datadist='m1.ddist')
myHTML( OR( m1) )
```

The ANOVA below shows the *p*-values for parameter estimates. All are significant
at the 5% level, i.e. "strong evidence" of an effect.

```{r label=ATS-M1-anova, dependson="ATS-M1", results='asis'} 
myHTML(anova( m1 ) )
```

The bootstrap validation shown below indicates the bias-corrected indices
are not very different from the model indices.

```{r label=ATS-M1-validate, dependson="ATS-M1",results='asis'} 
s <- validate( m1, B=200 )
myHTML( s )
```

Model 0 has an AIC of `r I(round(AIC(m0)))`, while the Model 1 AIC is
`r I(round(AIC( m1 )))`,  which is substantially lower.  We can also perform a 
NHST for the reduction in deviance, which is strongly significant, as shown 
below.

```{r label=ATS-Model-Comparison, dependson=c("ATS-M0", "ATS-M1"), results='asis'}
myHTML(lrtest( m0, m1 ))
```

```{r label=ATS-M1-confusion, dependson="ATS-M1"} 
m1.gof <- LR.summary( m1, print = FALSE )
```

Model 1 has much higer area under ROC curve (`r I(m1.gof$ROC)`%) and 
Nagelkerke *R^2* (`r  I(m1.gof$Nagelkerke)`%) than the null model.
The balanced accuracy, sensitivity, specificity of model 1 are
`r I(m1.gof$Accuracy)`, `r I(m1.gof$Sensitivity)` and 
`r I(m1.gof$Specificity)`. These indices are evidence of outstanding goodness 
of fit, discrimination and predictive accuracy.

### Model 2
Model 1 is essentially a replication of the OSSLS model previously published in 
the *Journal of Science and Medicine in Sport*, but slightly different because 
some variables are not present in both data sets.

To develop the model further, all the variables in the **Walking to School** and
**Walking  and Cycling to School**, **Travel in General**, **Physical Activity** 
and **Nutrition** sections of the questionnaire were added to Model 1, one at a 
time. This process resulted in two extra variables being added,  namely the 
degree to which the student feels in control of the travel mode decision and 
being a regular walker for general mobility. The latter variable has a low to 
moderate correlation with the number of cars available to the household 
(the Spearman rank correlation coefficient is
`r I(round(cor(as.numeric(dat$HMcars), as.numeric(dat$TgWalk), use='p', met='sp'),2))`), 
so including it alters the coefficient of cars, but the effects of each can 
still be intepreted independently.


```{r label=ATS-M2, dependson="ATS-M1", results='hide', cache=TRUE} 
options(datadist="m1.ddist")
m2 <- lrm( ATS_f ~ rcs(Dist2School,3) + n_cars + tired + onway + confd + 
             pshould + control + regwalk, x=T, y=T, data=m1.dat )
```

Model 2 has slightly higher GOF indices, and a significantly lower deviance than 
Model 1.
```{r label=ATS-M2-print, dependson="ATS-M2", results='asis'} 
m2.gof <- LR.summary( m2, html = TRUE )
```

The LR test indicates a significantly lower deviance, and the increase in GOF
is substantively pleasing also (e.g. ROCs of `r I(m1.gof$ROC)` for model 1  vs. 
`r I(m2.gof$ROC)` for model 2).

```{r label=ATS-M2-M1, dependson="ATS-M2", results='asis'} 
myHTML(lrtest(m1,m2) )
```

Odds ratios are shown below.
```{r label=ATS-M2-OR, dependson="ATS-M2", results='asis'} 
myHTML( OR( m2 ))
```

Models 1 & 2 meet the criterion of independent effects, i.e. the odds ratios 
for the variables common to both models are very similar, and well within the
95% confidence limits for each. Futhermore, the maximum VIF of model 2 is
`r require(rms); I(max(rms::vif(m2)))`, providing further evidence that the 
*ceteris paribus*  interpretation of the coefficients is valid.

```{r OR-comparisons, results='asis', dependson=c("ATS-M1", "ATS-M2")}
options(datadist="m1.ddist")
m1.OR <- OR( m1 )$OR
m2.OR <- OR( m2 )$OR
m2.OR <- m2.OR[ -c(6:7) ]
OR.names <- rownames(OR(m1))
myHTML( data.frame(Variable=OR.names,'OR (m1)'=m1.OR, 'OR (m2)'=m2.OR, 'Difference'=(m2.OR - m1.OR), check.names=F ) )
```

The figure below show the relationship of the log odds of ATS to each of the 
correlates in model 2, with shaded areas indicating 95% confidence intervals.

```{r m2-plot, dependson=c("ATS-M2", "data-setup")}

plot(Predict(m2))
```

Model 2 will be regarded as the final model for interpretation, unless there is 
a compelling reason to alter it. Statistically, it is sound. However one may 
wish to consider adding or replacing some variables on substantive grounds. 

As a final comment, one may think that because distance has such a large 
influence on ATS (The "null" model has an ROC of `r I(round(m0.gof$ROC))`%), 
that the additional variables in the final model do not have much explanatory
power in relative terms.  However this is not true.  Removing distance to school
from the final model still gives an ROC value of 
`r I(round(LR.summary(update(m2, . ~ . - rcs(Dist2School,3)))$ROC))`%, which is 
still outstanding.

## Interpretation of final model

In summary, the final model has absolutely outstanding predictive and 
discriminant validity, and all effects can be interpreted *ceteris paribus*. It 
also has good face validity.  Obvious environmental and household factors (the 
distance to school and availability of private motorised transport) decrease the 
likelihood of ATS. Logistical factors (school being on the way to somewhere 
else, most probably work, or the school of a sibling) also decrease the 
likelihood of ATS. Being confident that one can walk to school, being a regular 
walker in general, and having parental support or encouragement to use ATS 
increase the likelihood of ATS.

Perhaps the only non-obvious factor is being too tired to walk to school. Almost
half (`r I(round(prop.table(table(dat$WS2tired))["Agree"]*100,1))`%) of students
agreed or strongly agreed with this question. This suggests that parental factors
of:

- encouragement to use ATS
- strongly encouraging sleep and exercise patterns of behaviour that prevent 
  exessive tiredness in the morning;
- encouraging walking for general mobility

will be the most influential factors in raising  ATS rates among New Zealand 
secondary school students. 

These are, of course, components of a generally healthy lifestyle, However it 
may be relevant to note that there were many other healthy lifestyle factors 
that were unrelated to ATS. No variables in the nutrition or physical activity 
sections of the questionnaire influence the likelihood of ATS, for example.

## Colophon

All analyses were performed using R (ver. `r r_ver`)[R Core Team, 2013] and 
packages rms (ver. `r info$otherPkgs$rms$Version`) [F. Harrell, 2014] for 
analysis, Gmisc for plot and table output (ver. `r info$otherPkgs$Gmisc$Version`), 
and knitr (ver `r info$otherPkgs$knitr$Version`) [Xie, 2013] for reproducible 
research.
