---
title: "Correlates of Walking to School"
author: "John Williams"
output: 
  html_document:
    toc: true
    number_sections: true
    theme: cerulean
    css: Scripts/style.css
  Grmd::docx_document:
    fig_caption: TRUE
    force_captions: TRUE
    toc: TRUE
    toc_depth: 5
    css: Scripts/style.css 
  pdf_document:
    toc: true
    latex_engine: xelatex
    keep_tex: true
bibliography: /home/john/Dropbox/Writing/bib/all-refs.bib
csl: /home/john/Dropbox/Writing/bib/CSL/apa.csl

---

```{r label=R-setup, echo=FALSE, include=FALSE, cache=FALSE}
# “default”, “cerulean”, “journal”, “flatly”, “readable”, “spacelab”, “united”, “cosmo”
# bibliography: /home/john/Dropbox/Writing/bib/all-refs.bib
require( ggplot2 )
require( scales )
require( Gmisc )
require( rms )
require( knitr )
require( JMisc )
# Evaluate the figure caption after the plot
knitr::opts_knit$set(eval.after='fig.cap')

# Use the table counter that the htmlTable() provides
options(table_counter = TRUE)

# Use the figCapNo() with roman letters
options(fig_caption_no_roman = TRUE)

#theme_set( theme_gray( base_size = 10 ))
#theme_update( legend.key.width = unit( 3,"line") )
options(width=120)
options("show.signif.stars"=F)

## set global chunk options
opts_chunk$set(echo=FALSE,
               cache=FALSE,
               dpi=96,
               fig.height=5,
               fig.width=7,
               prompt=F,
               tidy=T,
               highlight=T,
               dev="png",
               dev.args=list(type="cairo"),
               fig.align='center',
               fig.show='hold',
               par=TRUE,
               comment=NA,
               background="wheat",
               prompt=FALSE,
               warning=FALSE,
               message=FALSE)

info   <- sessionInfo()
r_ver  <- paste(info$R.version$major, info$R.version$minor, sep=".")
barcol <- "wheat"
require(JMisc)
```


# Data description

## Data file
The **R** code below is just to show which data file is being used.

```{r label="read-data", echo=2:4, size="small", prompt=FALSE}
require(foreign)
d  <- "/home/john/Dropbox/Research/Collaboration/BEATS/John/Analysis/Correlates of ATS/"
f <- "BEATS_SS_ForWalk2School_160216_COMPLETEwithGIS.sav"
dat <- read.spss(fname(f, d), to.data.frame = TRUE, use.value.labels = FALSE)
rm(d, f)
```

```{r label=data-setup, results='hide'}
source("Scripts/data-setup-W2S.R")
```

## Variables

```{r mt, results='asis'}
cat("Name               Label\n-----------------  -----\n")
for(v in ats_vars)
   cat(sprintf("%-17s  %20s\n", v, getLab(dat, v)))
```


# Modeling

1. All cases in the file with a value of the variable **Include** equal 
   to 1 were included.  
1. Only cases where the distance to school (GIS was less than or equal to 2km 
   were included
1. Every potential covariate on the list that Sandy supplied was ran as a 
   "univariate" model in the sense above (i.e. the variable in question, plus
   the "moderators")
1. All the covariates that were significant at the 5% level were entered in a 
   multivariate model.
1. This procedure resulted in the models below. Given the relatively large 
   sample size, large number of IVs and large number of hypothesis tests, I 
   strongly recommend using the 1% level to judge statistical signficance.
1. The effect of distance on the probability of walking to school is clearly 
   non-linear, so the distance variable was transformed using a restricted cubic 
   spline (with three knots).
1. Because the data were collected within schools, robust standard errors were 
   calculated using school as a cluster variable. 

The **R** code below is to show which variables are included in which block, 
which correspond to elements of the Panter model.

- **EF**: External factors
- **PF**: Parental factors
- **YF**: Youth factos
- **PE**: Physical environment

```{r vars-setup, echo=TRUE}
mods <- "ATS_f ~ rcs(Dist2School_GIS, 3) + gender + Age_at_Survey" 
EF <- "+ WSbweather"
PF <- "+ HMcars + WSpsh + WSpunsafe + WSpwalk"
YF <- "+ WSAint + WSAnice + WSAstim + WSAgood + WSAuseful + WSAsafe + WCSone + WSchat + WSfsh + WSbfri + WSbcool + WSf5ws + WSbwant + WStired + WSno + WSbstuff + WSbsweat + WSbplan + WSbsched"
PE <- "+ F3_landuseaccess + WSbdist + WStime + WSbsafe + WSbfootp + WCSlights + WCStraffic + WCShills + WCSrbor"
```

## Model 0 
Here are the results of including all the significant "univariate" correlates,
plus the non-significant "moderators".  

**Note**: Dist2School' is the non-linear component of distance to school.

```{r m0, echo=TRUE}
m0.dat <- dat.ats[complete.cases(dat.ats$school, dat.ats$Dist2School_GIS) & dat.ats$Dist2School_GIS > 2000 & dat$Dist2School_GIS <= 4000,]
m.ddist <- datadist(m0.dat )
options( datadist='m.ddist' )
(m0 <- robcov(lrm( as.formula(paste(mods, EF, PE, PF, YF )), x=T, y=T, data=m0.dat ), cluster=m0.dat$school))
```
Because **Dlikedriven**  and **DParents** are not significant and including them
reduces the sample size considerably, I have removed them from all models.

## Model 1 
- Main moderators
- External factors
- Physical environment characteristics



```{r m1}
(m1 <- robcov(lrm(as.formula(paste(mods, EF, PE)), x=T, y=T, data=m0.dat ), cluster=m0.dat$school ))
```
  
## Model 2 

- Main moderators
- External factors
- Physical environment characteristics 
- Parental characteristics 
- Parental attitudes

```{r m2}
(m2 <- robcov(lrm(as.formula(paste(mods, EF, PE, PF)), x=T, y=T, data=m0.dat ), cluster=m0.dat$school))
```

## Model 3
(Same as Model 0)

- Main moderators
- External factors
- Physical environment characteristics 
- Parental characteristics 
- Parental attitudes
- Youth characteristics
- Youth attitudes

```{r m3}
(m3 <- robcov(lrm(as.formula(paste(mods, EF, PE, PF, YF)), x=T, y=T, data=m0.dat ), cluster=m0.dat$school))
```



## Model 4

To remove the non-significant variables, we can employ an automatic backward 
stepwise procedure.  I don't necessarily recommend this procedure, it's just
here for an example.

```{r bw, echo=TRUE}
fastbw(update(m3))
```


